```{r include=FALSE}
library(tidyverse)
library(car)
options(scipen=1)
```

# Complex surveys

**(Work in progress!)**

By the end of this chapter you will know how to:

* Setup a survey object using complex survey information such as sampling weight and stratification variables.
* Use a tidyverse-esq approach for descriptive statistics.
* Fit a GLM (logistic).

We will use the `survey` package and a wonderful tidyverse-style wrapper called `srvyr`.

```{r message=FALSE}
library(survey)
library(srvyr)
```


See [`srvyr` compared to the `survey` package](https://cran.r-project.org/web/packages/srvyr/vignettes/srvyr-vs-survey.html) and Fox and Weisberg's online appendix, [Fitting Regression Models to Data From Complex Surveys](https://socialsciences.mcmaster.ca/jfox/Books/Companion/appendices/Appendix-Surveys.pdf). The main reference for `survey` is Lumley (2010).



## The dataset

This chapter's dataset is drawn from the 2011 Canadian National Election Study -- taken from the `carData` package and described by the Fox and Weisberg appendix above. [Download it here](./ces11.csv).

There are 2231 observations on the following 9 variables:

| Variable name | Description |
|---------------|-------------|
| id            | Household ID number. |
| province      | a factor with (alphabetical) levels `AB`, `BC`, `MB`, `NB`, `NL`, `NS`, `ON`, `PE`, `QC`, `SK`; the sample was stratified by province. |
| population    | population of the respondent's province, number over age 17. |
| weight        | weight sample to size of population, taking into account unequal sampling probabilities by province and household size. |
| gender        | a factor with levels `Female`, `Male`. |
| abortion      | attitude toward abortion, a factor with levels `No`, `Yes`; answer to the question "Should abortion be banned?" |
| importance    | importance of religion, a factor with (alphabetical) levels `not`, `notvery`, `somewhat`, `very`; answer to the question, "In your life, would you say that religion is very important, somewhat important, not very important, or not important at all?" |
| education     | a factor with (alphabetical) levels `bachelors` (Bachelors degree), `college` (community college or technical school), `higher` (graduate degree), `HS` (high-school graduate), `lessHS` (less than high-school graduate), `somePS` (some post-secondary). |
| urban | place of residence, a factor with levels rural, urban. |


Read in the data.

```{r}
ces <- read.csv("ces11.csv", stringsAsFactors = TRUE)
head(ces)
```

I'm setting `stringsAsFactors` to `TRUE` so that the variables which are obviously factors are setup accordingly (R used to this by default; sometimes it had irritating side-effects).


## The components of a survey design

The key parts of the dataset which describe the survey design are as follows:

```{r}
ces %>% select(id, province, population, weight) %>%
  head(6)
```

* `id` is a unique identifier for each individual, which is particularly important when there is more than one data point per person, e.g., for multilevel modelling (not in this dataset).
* `province` the data were stratified by province -- random sampling by landline numbers was done within province
* `population` provides the population by province
* `weight` is the sampling weight, in this dataset calculated based on differences in province population, the study sample size therein, and household size

Here is how to setup a survey object using `srvyr`:

```{r}
ces_s <- ces %>%
  as_survey(ids = id,
            strata = province,
            fpc = population,
            weights = weight)
```



## Describing the data

We will sometimes want to compare weighted and unweighted analyses. Here is a warmup activity to show how using the tidyverse.


### Activity

a. use the `ces` data frame and `tidyverse` to calculate the number of people who think abortion should be banned
b. do the same again, but this time use the `ces_s` survey object created above -- what do you notice?
c. compare the proportions saying `yes` by group

**Hint:** you will want to use `group_by`.

**Another hint:** to count, use the function `n`. The version for survey objects is called `survey_total`.


### Answer

**a. use the `ces` data frame and `tidyverse` to calculate the number of people who think abortion should be banned**

```{r}
ces %>%
  group_by(abortion) %>%
  summarise(n = n())
```

**b. do the same again, but this time use the `ces_s` survey object created above -- what do you notice?**

```{r}
ces_s %>%
  group_by(abortion) %>%
  summarise(n = survey_total())
```


The counts are much bigger than the number of rows in the dataset due to the sampling weights.


**c. compare the proportions saying `yes` by group**

One way to do this is by copy and paste!

Unweighted:

```{r}
prop_unweighted <- 413 / (413 + 1818)
prop_unweighted
```

Weighted:

```{r}
prop_weighted <- 2964018 / (2964018 +	13059520)
prop_weighted
```

The unweighted proportion of "yes" is only a little different in this case: 0.1851188 (unweighted) v 0.184979 (weighted), though that represents a population difference of over 2,000 people.

Here's how to answer the questions in one go:

```{r}
ces_s %>%
  group_by(abortion) %>%
  summarise(n_raw = unweighted(n()),
            n_weighted = survey_total()) %>%
  mutate(prop_raw = n_raw / sum(n_raw),
         prop_weighted = n_weighted / sum(n_weighted)) %>%
  select(-n_weighted_se)
```


## Fitting a GLM

The `survey` package makes this very easy. There is a command called `svyglm` which is identical to `glm` except it has parameter called `design` instead if `data`.

See `?svyglm`


### Activity

a. mutate the survey object to add a binary variable called `againstAbortion` which is 1 if the participant is against abortion and 0 if not.
b. fit an intercept-only logistic regression model without using weights (you can use `as_tibble` to get the "raw" data frame hidden within the survey object).
c. Do the same again, this time using the survey structure.
d. compare the predicted proportions with the "raw" proportions we calculated earlier

### Answer


**a. mutate the survey object to add a binary variable called `againstAbortion` which is 1 if the participant is against abortion and 0 if not.**

```{r}
ces_s <- ces_s %>%
  mutate(againstAbortion = as.numeric(ces$abortion == "Yes"))
```

**b. fit an intercept-only logistic regression model without using weights (you can use `as_tibble` to get the "raw" data frame hidden within the survey object).**

```{r}
m0 <- glm(againstAbortion ~ 1,
          data = as_tibble(ces_s),
          family = binomial)
summary(m0)
```

**c. Do the same again, this time using the survey structure.**

```{r}
sm0 <- svyglm(againstAbortion ~ 1, design = ces_s,
              family = binomial)
summary(sm0)
```

**d. compare the predicted proportions with the "raw" proportions we calculated earlier**

This undoes the log-odds (logit) transform:

```{r}
exp(coef(m0)) / (1+exp(coef(m0)))
exp(coef(sm0)) / (1+exp(coef(sm0)))
```

The answer is the same as for the simple unweighted and weighted proportions, respectively.


## Slopes

Now, having completed the traditional step of fitting an intercept-only model, we can give the slopes a go.

The `Anova` command in the `car` package works for `svyglm` models as before.


### Activity

Regress `againstAbortion` on `importance`, `education`, and `gender`, and interpret what you find.

### Answer

```{r}
sm1 <- svyglm(againstAbortion ~ importance + education + gender,
              design = ces_s,
              family = binomial)
summary(sm1)
```


To interpret those categorical predictors, it will help to check what the levels are:

```{r}
levels(ces$importance)
```

So "not" is the comparison level.

```{r}
levels(ces$education) 
```

"bachelors" is the comparison level.

Gender is just a binary.

Example interpretations:

* Men were more likely to be against abortion (log-odds 0.33 more)

* People for whom religion was very important were more likely than those who said "not important at all" to be against abortion (log-odds 3.14)

You could get the odds ratios like so:

```{r}
exp(coef(sm1)) %>% round(2)
```

So the odds ratio of "very" v "not" important is 23.1.

The `ggeffects` package also works with `survey` models:

```{r}
library(ggeffects)
ggeffect(sm1, terms = c("education", "importance"))  %>%
  plot() +
  ylim(0,1)
```




## Diagnostics

Many of the diagnostic check we previously encountered work here too.

Here are the VIFs:

```{r}
vif(sm1)
```


My favorite, the DFBETA plots:

```{r fig.height=6, fig.width=6}
dfbetaPlots(sm1)
```


Try also:

```{r eval = FALSE}
influence.measures(sm1)
```




