```{r include=FALSE}
library(tidyverse)
library(car)
options(scipen=1)
```

# Mediation analysis

**(Work in progress!)**

In this example, let's make up some data for an experiment where people are randomised to one of two groups. The intervention group receives £2000 and the control group receives £500. The hypothesis is that one month later, the intervention group will have more savings than the control group; i.e., they won't just spend all of the money they have received.

```{r}
set.seed(42)
study_n <- 200
dat <- tibble(savings_start = rnorm(study_n, 5000, 2000) %>%
                ifelse(. < 0, 0, .),
              treat_group    = rbinom(study_n, size = 1, p = .5),
              money_received = ifelse(treat_group == 1,
                                      2000,
                                      500) + rnorm(study_n, 0, 5),
              savings_end    = -500 + savings_start + money_received +
                                 rnorm(study_n, 0, 600) %>%
                ifelse(. < 0, 0, .))
                            
dat <- dat %>%
  mutate(treat_group = dplyr::recode(treat_group,
                                       `1` = "Intervention",
                                       `0` = "Control")) %>%
  mutate(treat_group = as.factor(treat_group))
dat
```

Here's a graph:

```{r fig.height=3, fig.width=4}
dat %>%
  group_by(treat_group) %>%
  summarise(savings_Pre  = mean(savings_start),
            savings_Post = mean(savings_end)) %>%
  pivot_longer(cols = starts_with("savings"),
               names_to = "time",
               values_to = "savings",
               names_prefix = "savings_") %>%
  mutate(time = factor(time)) %>%
  mutate(time = relevel(time, "Pre")) %>%
  ggplot(aes(x = time, y = savings, fill = treat_group)) +
    geom_col(position = "dodge") +
    labs(x = "Time",
         y = "Savings",
         fill = "Group")
```

Fit the two models:

```{r}
mod_med <- lm(money_received ~ savings_start + treat_group, data = dat)
mod_out <- lm(savings_end ~ savings_start + treat_group + money_received, data = dat)
```

Have a look:

```{r}
summary(mod_med)
```

The coefficients for the model regressing savings on treatment group and money received wipe each other out...

```{r}
summary(mod_out)
```

Check out the VIF:

```{r}
vif(mod_out)
```

HUGE, as expected.

Test the mediation:

```{r}
library(mediation)
test_med <- mediate(model.m = mod_med,
                    model.y = mod_out,
                    treat = "treat_group",
                    mediator = "money_received",
                    robustSE = TRUE,
                    sims = 100,
                    control.value = "Control",
                    treat.value = "Intervention")
summary(test_med)
```
