```{r include=FALSE}
library(tidyverse)
library(car)
options(scipen=1)
```

# Complex surveys

By the end of this chapter you will know how to:

* Setup a survey object using complex survey information such as sampling weight and stratification variables.
* Use a tidyverse-esq approach for descriptive statistics.
* Fit a GLM (logistic).

We will use the `survey` package and a tidyverse-style wrapper called `srvyr`.

```{r message=FALSE}
library(survey)
library(srvyr)
```


## Readings

These are handy:

* [`srvyr` compared to the `survey` package](https://cran.r-project.org/web/packages/srvyr/vignettes/srvyr-vs-survey.html) explains a way to use survey data in the tidyverse.
* Fox and Weisberg's online appendix, [Fitting Regression Models to Data From Complex Surveys](https://socialsciences.mcmaster.ca/jfox/Books/Companion/appendices/Appendix-Surveys.pdf).
* The main reference for the models implemented by `survey` is the (expensive) book by Lumley (2010).
* UCLA has [extensive notes](https://stats.idre.ucla.edu/r/seminars/survey-data-analysis-with-r/) from a 2020 seminar on survey analysis.
* [Analyzing international survey data with the pewmethods R package](https://medium.com/pew-research-center-decoded/analyzing-international-survey-data-with-the-pewmethods-r-package-3b0b21cba607), by Kat Devlin, explains an alternative way to use weights for descriptive stats.



## The dataset

This chapter's dataset is drawn from the 2011 Canadian National Election Study -- taken from the `carData` package and described by the Fox and Weisberg appendix cited above. [Download it here](./ces11.csv).

There are 2231 observations on the following 9 variables:

| Variable name | Description |
|---------------|-------------|
| id            | Household ID number. |
| province      | a factor with (alphabetical) levels `AB`, `BC`, `MB`, `NB`, `NL`, `NS`, `ON`, `PE`, `QC`, `SK`; the sample was stratified by province. |
| population    | population of the respondent's province, number over age 17. |
| weight        | weight sample to size of population, taking into account unequal sampling probabilities by province and household size. |
| gender        | a factor with levels `Female`, `Male`. |
| abortion      | attitude toward abortion, a factor with levels `No`, `Yes`; answer to the question "Should abortion be banned?" |
| importance    | importance of religion, a factor with (alphabetical) levels `not`, `notvery`, `somewhat`, `very`; answer to the question, "In your life, would you say that religion is very important, somewhat important, not very important, or not important at all?" |
| education     | a factor with (alphabetical) levels `bachelors` (Bachelors degree), `college` (community college or technical school), `higher` (graduate degree), `HS` (high-school graduate), `lessHS` (less than high-school graduate), `somePS` (some post-secondary). |
| urban | place of residence, a factor with levels rural, urban. |


Read in the data.

```{r}
ces <- read.csv("ces11.csv", stringsAsFactors = TRUE)
```

I'm setting `stringsAsFactors` to `TRUE` so that the variables which are obviously factors are setup accordingly (R used to this by default; sometimes it had irritating side-effects).


## The components of a survey design

The key parts of the dataset which describe the survey design are as follows:

```{r}
ces %>% select(id, province, population, weight) %>%
  head(6)
```

* `id` is a unique identifier for each individual, which is particularly important when there is more than one data point per person, e.g., for multilevel modelling (not in this dataset).
* `province` the data were stratified by province -- random sampling by landline numbers was done within province.
* `population` provides the population by province.
* `weight` is the sampling weight, in this dataset calculated based on differences in province population, the study sample size therein, and household size.

Here is how to setup a survey object using `srvyr`:

```{r}
ces_s <- ces %>%
  as_survey(ids = id,
            strata = province,
            fpc = population,
            weights = weight)
```



## Describing the data

We will sometimes want to compare weighted and unweighted analyses. Here is a warmup activity to show how using the tidyverse.


### Activity

a. use the `ces` data frame and `tidyverse` to calculate the number of people who think abortion should be banned
b. do the same again, but this time use the `ces_s` survey object created above -- what do you notice?
c. compare the proportions saying `yes` by group

**Hint:** you will want to use `group_by`.

**Another hint:** to count, use the function `n`. The version for survey objects is called `survey_total`.


### Answer

**a. use the `ces` data frame and `tidyverse` to calculate the number of people who think abortion should be banned**

```{r}
ces %>%
  group_by(abortion) %>%
  summarise(n = n())
```

**b. do the same again, but this time use the `ces_s` survey object created above -- what do you notice?**

```{r}
ces_s %>%
  group_by(abortion) %>%
  summarise(n = survey_total())
```


The counts are much bigger than the number of rows in the dataset due to the sampling weights.


**c. compare the proportions saying `yes` by group**

One way to do this is by copy and paste!

Unweighted:

```{r}
prop_unweighted <- 413 / (413 + 1818)
prop_unweighted
```

Weighted:

```{r}
prop_weighted <- 2964018 / (2964018 +	13059520)
prop_weighted
```

The unweighted proportion of "yes" is only a little different in this case: 0.1851188 (unweighted) v 0.184979 (weighted).

Here's how to answer the questions in one go:

```{r}
ces_s %>%
  group_by(abortion) %>%
  summarise(n_raw = unweighted(n()),
            n_weighted = survey_total()) %>%
  mutate(prop_raw = n_raw / sum(n_raw),
         prop_weighted = n_weighted / sum(n_weighted)) %>%
  select(-n_weighted_se)
```


## Fitting a GLM

The `survey` package makes this very easy. There is a command called `svyglm` which is identical to `glm` except it has parameter called `design` instead of `data`.

See `?svyglm`


### Activity

a. `mutate` the survey object to add a binary variable called `againstAbortion` which is 1 if the participant is against abortion and 0 if not.
b. fit an intercept-only logistic regression model without using weights (you can use `as_tibble` to get the "raw" data frame hidden within the survey object).
c. Do the same again, this time using the survey structure.
d. compare the predicted proportions with the "raw" proportions we calculated earlier

### Answer


**a. `mutate` the survey object to add a binary variable called `againstAbortion` which is 1 if the participant is against abortion and 0 if not.**

```{r}
ces_s <- ces_s %>%
  mutate(againstAbortion = as.numeric(ces$abortion == "Yes"))
```

**b. fit an intercept-only logistic regression model without using weights (you can use `as_tibble` to get the "raw" data frame hidden within the survey object).**

```{r}
m0 <- glm(againstAbortion ~ 1,
          data = as_tibble(ces_s),
          family = binomial)
summary(m0)
```

**c. Do the same again, this time using the survey structure.**

```{r}
sm0 <- svyglm(againstAbortion ~ 1, design = ces_s,
              family = binomial)
summary(sm0)
```

**d. compare the predicted proportions with the "raw" proportions we calculated earlier**

This undoes the log-odds (logit) transform:

```{r}
exp(coef(m0)) / (1+exp(coef(m0)))
exp(coef(sm0)) / (1+exp(coef(sm0)))
```

The answer is the same as for the simple unweighted and weighted proportions, respectively.


## Slopes

Now, having completed the traditional step of fitting an intercept-only model, we can give the slopes a go.

The `Anova` command in the `car` package works for `svyglm` models as before.


### Activity

Regress `againstAbortion` on `importance`, `education`, and `gender`, and interpret what you find.

### Answer

```{r}
sm1 <- svyglm(againstAbortion ~ importance + education + gender,
              design = ces_s,
              family = binomial)
summary(sm1)
```


To interpret those categorical predictors, it will help to check what the levels are:

```{r}
levels(ces$importance)
```

So "not" is the comparison level.

```{r}
levels(ces$education) 
```

"bachelors" is the comparison level.

Gender is just a binary.

Example interpretations:

* Men were more likely to be against abortion (log-odds 0.33 more)

* People for whom religion was very important were more likely than those who said "not important at all" to be against abortion (log-odds 3.14)

You could get the odds ratios like so:

```{r}
exp(coef(sm1)) %>% round(2)
```

So the odds ratio of "very" versus "not" important is 23.1.

The `ggeffects` package also works with `survey` models (hurrah):

```{r message=FALSE}
library(ggeffects)
ggeffect(sm1, terms = c("education", "importance"))  %>%
  plot() +
  ylim(0,1)
```




## Diagnostics

Many of the diagnostic checks we previously encountered work here too.

Here are the VIFs:

```{r}
vif(sm1)
```


My favorite, the DFBETA plots:

```{r fig.height=6, fig.width=6}
dfbetaPlots(sm1)
```


Try also:

```{r eval = FALSE}
influence.measures(sm1)
```



## Another worked example: the European Social Survey

NatCen recently [published](https://www.bsa.natcen.ac.uk/latest-report/british-social-attitudes-37/fairness-and-justice-in-britain.aspx) analyses of views on fairness and justice in Britain (Curtice, Hudson, & Montagu, 2020), using data from the ninth wave of the 2019 European Social Survey (European Social Survey Round 9 Data, 2018).

Let's see if we can replicate the results.

Firstly, you will have to download the data from [the ESS website](https://www.europeansocialsurvey.org/data/download.html?r=9). Go for the SPSS version. (You will have to register but it is a very quick process.) It will arrive as a zip file (`ESS9e02.spss.zip`). Simply extract the SPSS file therein (`ESS9e02.sav`) and move it to the same folder as your Markdown file.

```{r}
library(haven)
ess9 <- read_sav("ESS9e02.sav")
```

Try having a look:

```{r eval = FALSE}
View(ess9)
```

Okay, if that has worked then the next problem is finding the variables we want to analyse!

The report opens with a spotlight figure:

> Only 20% of the British public think that differences in wealth in Britain are fair, whilst a majority (59%) think that wealth differences in Britain are unfairly large and a further 16% think that differences in wealth are unfairly small.

The question asked was:

> In your opinion, are differences in wealth in Britain unfairly small, fair, or unfairly large? 

It's a good exercise to spend some time wading through the documentation on the ESS website to find the variable name, before looking at the answer below...

We will also need the variable for country (easier to spot) and any information required for setting up the survey object. The ESS website helpfully advises:

> In general, you must weight tables before quoting percentages from them. The Design weights (DWEIGHT) adjust for different selection probabilities, while the Post-stratification weights (PSPWGHT) adjust for sampling error and non-response bias as well as different selection probabilities. Either DWEIGHT or PSPWGHT must always be used. In addition, the Population size weights (PWEIGHT) should be applied if you are looking at aggregates or averages for two or more countries combined. See the guide Weighting European Social Survey Data for fuller details about which weights to use.

This also links to a guide to [weighting the data](https://www.europeansocialsurvey.org/docs/methodology/ESS_weighting_data_1_1.pdf).

It's also worth printing all the variable names -- if only to spot that they have ended up in lower case.


### Set up the survey object

First, I'm going to fix the country variable. It currently looks like:

```{r}
table(ess9$cntry)
```

But the dataset also has nicer labels included, which we can get like this using the `as_factor` function (note the underscore). This function is in the `haven` package.


```{r}
ess9$cntry <- as_factor(ess9$cntry, levels = "labels")
table(ess9$cntry)
```

Now let's setup the survey object:

```{r}
ess9_survey <- ess9 %>%
  as_survey_design(ids = idno,
                   strata = cntry,
                   nest = TRUE,
                   weights = pspwght)
```

The `nest` option takes account of the `ids` being nested within strata: in other words the same ID is used more than once across the dataset but only once in a country.


### Try the analysis

The country variable is `cntry` and the wealth variable is `wltdffr`, which I spotted with the help of the [code book](https://www.europeansocialsurvey.org/docs/round9/survey/ESS9_appendix_a7_e02_0.pdf).

The first thing you will spot is that the original variable is coded from -4 to 4: 

| Code | Meaning |
|---|-------------------------|
|-4 | Small, extremely unfair |
|-3 | Small, very unfair |
|-2 | Small, somewhat unfair |
|-1 | Small, slightly unfair |
|0  | Fair |
|1  | Large, slightly unfair |
|2  | Large, somewhat unfair |
|3  | Large, very unfair |
|4  | Large, extremely unfair |

So let's create another variable which is grouped as per the NatCen report:

```{r}
ess9_survey <- ess9_survey %>%
  mutate(wltdffr_group =
           case_when(
             wltdffr >= -4 &
               wltdffr <= -1 ~ "Unfairly small",
             wltdffr == 0 ~ "Fair",
             wltdffr >= 1 & wltdffr <= 4 ~ "Unfairly large"
           ),
         wltdffr_group = factor(wltdffr_group,
                                  levels = c("Unfairly small",
                                             "Fair",
                                             "Unfairly large"))
  )
```


```{r}
gb_wealth <- ess9_survey %>%
  filter(cntry == "United Kingdom") %>%
  group_by(wltdffr_group) %>%
  summarise(prop = survey_mean(vartype = "ci"))
gb_wealth 
```

The results are the same as per the report. Let's round to show this more clearly:

```{r}
gb_wealth %>%
  mutate(perc = (prop*100) %>% round(0)) %>%
  select(wltdffr_group, perc)
```

We can also plot the results:

```{r fig.height=6, fig.width=5}
gb_wealth %>%
  filter(!is.na(wltdffr_group)) %>%
  ggplot(aes(x = wltdffr_group, y = prop*100)) +
  geom_col(fill = "#B053A1") +
  geom_errorbar(aes(ymin = prop_low*100,
                    ymax = prop_upp*100), width = 0.2) +
  ylim(0,100) +
  labs(y = "%", x = NULL,
       title = "In your opinion, are differences in wealth in Britain\nunfairly small, fair, or unfairly large?")
```


Let's do it again for a selection of countries. First, make a function which carries out the analysis for one country:

```{r}
gimme_country_results <- function(the_cntry) {
  ess9_survey %>%
    filter(cntry == the_cntry) %>%
  group_by(wltdffr_group) %>%
  summarise(prop = survey_mean(vartype = "ci")) %>%
  mutate(cntry = the_cntry)
}
```

Check it works for the UK:

```{r}
gimme_country_results("United Kingdom")
```

Run it for all the countries of interest:

```{r}
conts <- c("Germany", "Spain", "France", "United Kingdom", "Italy")
euro_wealth <- map_dfr(conts, gimme_country_results)
head(euro_wealth)
```


Next, try a plot:

```{r fig.height=5, fig.width=6}
euro_wealth %>%
  filter(!is.na(wltdffr_group)) %>%
  ggplot(aes(x = cntry,
             y = prop*100,
             ymin = prop_low*100,
             ymax = prop_upp*100,
             fill = wltdffr_group)) +
  geom_col(position = position_dodge(width = .8), width = 0.6) + 
  geom_errorbar(position=position_dodge(width = .8),
                colour="black",
                width = 0.2) +
  ylim(0,100) +
  labs(y = "%", x = NULL,
       title = "In your opinion, are differences in wealth\nunfairly small, fair, or unfairly large?",
       fill = NULL)
```
